\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-05-18 manuscript_v5/02-methods.qmd
discussion, appendix updated

We used data from the *Survey of COVID-19 related Behaviours and Attitudes*, a repeated cross sectional survey focused on the Canadian province of Ontario that was commissioned by the Fields Institute for Research in Mathematical Sciences and the  Mathematical Modelling of COVID-19 Task Force under ethical guidance from the University of Toronto, and which ran between September 30th, 2021 and January 17th,2022. The survey collected socio-economic information from participants (@tbl-descriptive-stats), their location (nearest municipality,  as shown in @fig-map), the date of access to the survey, and asked information on vaccination status by using the question "Have you received the first dose of the COVID vaccine?", with possible answers "yes" and "no". The original dataset contained 39,029 observations (where each observation corresponded to a unique respondent).

Preliminary analyses of the data included the removal of outliers (**should we still do this? it's only 19 observations with income >110k and household of 1, but we are not even using such income bracket in the analysis because we re-grouped the data, and the household size variable has 90% missing rate**), of observations where respondents did not provide answers in all the covariates of interest, matching the city of each observations with its corresponding LHIN and Health Region, and removing observations from areas with low representation (107 observations corresponding to the North West and North East Health Regions). After all the preliminary analyses indicated above, the total number of observations used for analysis was 3,549 which included the East, Central, Toronto, and West Health Regions covering the period between October 1st,2021 and December 12th, 2021. The original dataset, clean dataset, and details on the data cleaning process are described in detail in the [GitHub repository](https://github.com/aimundo/Fields_COVID-19/) for this paper.

```{r,map}
#| label: fig-map
#| fig-cap: "Geographic representation of the data collected by the _Survey of COVID-19 related Behaviours and Attitudes_, collected by the Fields Institute in Ontario. The municipalities (cities) from where survey participants provided answers (in the clean dataset) appear as points. The Health six Regions are color-coded. Internal boundaries within certain Health Regions indicate areas that belonged to the Local Integrated Health Networks (LHINs), the geographic areas for healthcare in Ontario before the adoption of the Health Regions."
#| message: false
#| echo: false
#| fig-height: 3
#| figh-width: 2

>>>>

We used data from the *Survey of COVID-19 related Behaviours and Attitudes*, a repeated cross sectional survey focused on the Canadian province of Ontario that was commissioned by the Fields Institute for Research in Mathematical Sciences and the  Mathematical Modelling of COVID-19 Task Force under ethical guidance from the University of Toronto, and which ran between September 30th, 2021 and January 17th,2022. The survey collected socio-economic information from participants (@tbl-descriptive-stats), their location (nearest municipality,  as shown in @fig-map), the date of access to the survey, and asked information on vaccination status by using the question "Have you received the first dose of the COVID vaccine?", with possible answers "yes" and "no". The original dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and deployed the questions randomly, which resulted in $\approx$ 84\% of the observations having multiple missing answers or being incomplete.  Therefore, we selected 6,343 observations that were labeled as "complete" and that had complete answers for all covariates considered in our analysis. Later, we matched the city of each observation with its corresponding LHIN and Health Region, and removed observations from areas with low representation (107 observations corresponding to the North West and North East Health Regions). After all the preliminary analyses indicated above, the total number of observations used for analysis was 6,236 which included the East, Central, Toronto, and West Health Regions covering the period between October 1st,2021 and January 17, 2022. The original dataset, clean dataset, and details on the data cleaning process are described in detail in the [GitHub repository](https://github.com/aimundo/Fields_COVID-19/) for this paper.

```{r,map}
#| label: fig-map
#| fig-cap: "Geographic representation of the data collected by the _Survey of COVID-19 related Behaviours and Attitudes_, collected by the Fields Institute in Ontario. The municipalities (cities) from where survey participants provided answers (in the clean dataset) appear as points. The Health six Regions are color-coded. Internal boundaries within certain Health Regions indicate areas that belonged to the Local Integrated Health Networks (LHINs), the geographic areas for healthcare in Ontario before the adoption of the Health Regions."
#| message: false
#| echo: false
#| fig-height: 12
#| figh-width: 10

### 2023-05-18 manuscript_v5/03-results.qmd
discussion, appendix updated

@tbl-descriptive-stats shows the characteristics of the data from the Fields COVID-19 survey used for analysis. The sample contained **6,236** observations, from which 24.8\% (1,547) corresponded to individuals that reported not having received the first dose of the vaccine. Vaccination rates ranged between 71-79\% across household income brackets, age groups, Health Regions, and the months considered in the survey. However, the highest vaccination rates in each category were reported by individuals in the highest income bracket (79\%), those between 16 and 34 years of age (77\%), individuals that lived in the East Health Region (77\%), and  during January of 2022 (78\%). Differences were higher between racial/ethnic groups, where the higher vaccination rate was reported by White/Caucasian individuals (84\%), against vaccination rates between 63-66\% reported in the case of Arab/Middle Eastern, Black, Indigenous, Latin American individuals, and those that reported belonging to "Other" racial groups, which included Southeast Asian, Filipino, West Asian, and minorities not identified elsewhere.

@@

@fig-models presents the estimates (as odd ratios) from the logistic regression models for vaccination status using the socio-demographic factors collected by the survey, and their interactions. Generally speaking, lower odds of vaccination were identified in both cases in individuals characterized by a low household income, or that identified as part of equity-deserving groups. However, the magnitude of the estimates differed between the uncorrected and corrected models and more importantly, certain estimates were not deemed to be statistically-significant after the correction, in contrast to the estimates from the uncorrected model. Specifically, the corrected model showed no significant differences in vaccination odds between the age groups considered, the East Health Region, Latin American individuals with a household income under CAD 25,000, and Indigenous individuals living in the Central Health Region (@fig-models,B).
<!-- \newpage -->

@@

```{r,models}
#| message: false
#| warning: false
#| echo: false
#| fig-align: left
#| fig-width: 15
#| fig-height: 14
#| label: fig-models
#| fig-cap: "Coefficient estimates and confidence intervals for the uncorrected model. Only statistically significant interaction terms are shown. Full interaction terms can be found in Supplementary Figure A-3."

@@

Interestingly, individuals in equity-deserving groups with a household income below CAD 25,000 had higher odds of vaccination (when compared to those with a household income above CAD 60,000).  This held true in the case of Arab/Middle Eastern (OR=34, CI=[1.70,6.79]), Black individuals (OR=3.81, CI=[2.05, 7.09]), or those in other racial or ethnic groups (OR=3.19, CI=[2.00,5.09]). Additionally, individuals with an income between CAD 25,000 and 59,999 in the Arab/Middle Eastern and other racial ethnic groups had higher odds of vaccination (ORs and CIs=6.96 [2.67,18.16], and 3.5 [1.85,6.62]).

>>>>

@tbl-descriptive-stats shows the characteristics of the data from the Fields COVID-19 survey used for analysis. The sample contained 6,236 observations, from which 24.8\% (1,547) corresponded to individuals that reported not having received the first dose of the vaccine. Vaccination rates ranged between 71-79\% across all household income brackets, age groups, Health Regions, and the months considered in the survey. However, the highest vaccination rates in each category were reported by individuals in the highest income bracket (79\%), those between 16 and 34 years of age (77\%), individuals that lived in the East Health Region (77\%), and  during January of 2022 (78\%). Between racial/ethnic groups, the highest vaccination rate was reported by White/Caucasian individuals (84\%), against vaccination rates between 63-66\% reported in the case of Arab/Middle Eastern, Black, Indigenous, Latin American individuals, and those that reported belonging to "Other" racial groups, which included Southeast Asian, Filipino, West Asian, and minorities not identified elsewhere.

@@

@fig-models presents the estimates (as odd ratios) from the logistic regression models for vaccination status using the socio-demographic factors collected by the survey, and their interactions. Generally speaking, lower odds of vaccination were identified in both cases in individuals characterized by a low household income, or that identified as part of equity-deserving groups. However, the magnitude of the estimates differed between the uncorrected and corrected models and more importantly, there were differences in the statistical significance of certain estimates before and after the correction. Specifically, the uncorrected model showed significant differences in vaccination odds between the age groups considered, the East Health Region, Latin American individuals with a household income under CAD 25,000, and Indigenous individuals living in the Central Health Region (@fig-models,B) but these were deemed non statistically-significant after the correction.

<!-- \newpage -->

@@

```{r,models}
#| message: false
#| warning: false
#| echo: false
#| fig-align: left
#| fig-width: 15
#| fig-height: 14
#| label: fig-models
#| fig-cap: "Coefficient estimates and confidence intervals for the uncorrected model. Only statistically significant interaction terms are shown. Full interaction terms can be found in Supplementary Figures A-3 and A-4."

@@

Interestingly, individuals in equity-deserving groups with a household income below CAD 25,000 had higher odds of vaccination (when compared to those with a household income above CAD 60,000).  This held true in the case of Arab/Middle Eastern individuals (OR=34, CI=[1.70,6.79]), Black individuals (OR=3.81, CI=[2.05, 7.09]), and those in other racial or ethnic groups (OR=3.19, CI=[2.00,5.09]). Additionally, individuals with an income between CAD 25,000 and 59,999 in the Arab/Middle Eastern and other racial ethnic groups also had higher odds of vaccination than their high-income peers (ORs and CIs=6.96 [2.67,18.16], and 3.5 [1.85,6.62]).

### 2023-05-18 manuscript_v5/04-discussion.qmd
discussion, appendix updated

In this study we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions during between late 2021 and early 2022, aiming at determining which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensure that the Health Region model is able to fulfill its mission of improving health access for the inhabitants of Ontario.

Our results show that indeed, there were differences in vaccination odds across Ontario in certain socio-demographic groups. Specifically, those who identified as Arab/Middle Eastern, Black, Latin American, having mixed racial or ethnic background, or that belonged to other groups not explicitly included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were between a third and a half of that of individuals that identified as White/Caucasian (@fig-models). These results are consistent with previous studies that have shown lower vaccination rates in individuals with the same socio-demographic characteristics [@guay2022;@muhajarine2021;@carter2022;@hussain2022].

Lower vaccine uptake in the socio-demographic groups indicated above may be influenced in part, by vaccine hesitancy and refusal, which have been associated in equity-deserving Canadian individuals to concerns on vaccine safety, effectiveness, and experiences of racial discrimination in health settings [@basta2022;@cnat2022a;@cnat2022b;@cnat2023]. However, it has been shown that structural barriers also play an important role in vaccination uptake. In the case of equity-deserving individuals, such barriers include complex scheduling systems, language barriers, lack of adequate public transportation, and lack of accessible vaccination sites [@njoku2021]. In this regard, it is interesting to note that vaccination venues were scarce in low socio-economic areas that had the highest burden of COVID-19 in Toronto and other regions of Ontario around the time covered by the survey [@bogoch2022;@iveniuk2021], and that pharmacies in the Peel region (an area identified as a "hotspot" with high numbers of essential workers and multigenerational households) could not keep up with demand [@gill2022]. This suggests disparities in vaccine accessibility that affected in particular equity-deserving individuals in Ontario at the time of the survey. However, because to the best of our knowledge there seems to be a very limited amount of literature on this topic in the context of Ontario, there is an ongoing need of future studies that examine the longitudinal impact of vaccine accessibility and structural barriers that affect equity-deserving groups within the province.

@@

However, the results also indicate that the place of habitation affected the odds of vaccination for certain equity-deserving groups (interaction term of Health Region and Race, @fig-models). Specifically, this held true in the case of individuals identifying as Indigenous or with mixed racial background in the West Health Region, Black individuals in the Central Health Region, and Arab/Middle Eastern individuals in the East Health Region [@fig-models]. For these individuals, vaccination odds were lower when compared to the Toronto Health Region (Supplementary Figure A-6). We indicate next some contributing factors that might help provide context in each case.

First, it is useful to analyze the data using the LHINs, because most of the studies in the literature that have analyzed health in Ontario use the LHINs as the base of their analyses. Interestingly, for Indigenous and mixed individuals most of the observations in the survey for West Health Region came from the Hamilton Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs (**add table in Appendix**), whereas for Arab/Middle Eastern Individuals in the East Health Region the highest number of observations corresponded to the Champlain and Central East LHINs. Previous research has identified health disparities in these (mostly rural) regions, such as unequal distribution of primary care providers, increased mortality, and low pharmacist availability [@shah2019;@crighton2015;@timony2022].

>>>>

In this study we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions during between late 2021 and early 2022. Our goal was to determine which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensure that the Health Region model is able to fulfill its mission of improving health access for all Ontarians.

Our results show that indeed, there were differences in vaccination odds across Ontario in certain socio-demographic groups. Specifically, those who identified as Arab/Middle Eastern, Black, Latin American, having mixed racial or ethnic background, or that belonged to other groups not explicitly included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were between a third and a half of that of individuals that identified as White or Caucasian (@fig-models). These results are consistent with previous studies that have shown lower vaccination rates in individuals with the same socio-demographic characteristics [@guay2022;@muhajarine2021;@carter2022;@hussain2022].

Lower vaccine uptake in the socio-demographic groups indicated above may be influenced in part, by vaccine hesitancy and refusal, which have been associated in equity-deserving Canadian individuals to concerns on vaccine safety, effectiveness, and experiences of racial discrimination in health settings [@basta2022;@cnat2022a;@cnat2022b;@cnat2023]. However, it has been shown that structural barriers also play an important role in vaccination uptake. In the case of equity-deserving individuals, such barriers include complex scheduling systems, language barriers, lack of adequate public transportation, and lack of accessible vaccination sites [@njoku2021]. In this regard, it is interesting to note that vaccination venues were scarce in low socio-economic areas that had the highest burden of COVID-19 in Toronto and other regions of Ontario around the time covered by the survey [@bogoch2022;@iveniuk2021], and that pharmacies in the Peel region (an area identified as a "hotspot" with high numbers of essential workers and multigenerational households) could not keep up with demand [@gill2022]. This suggests disparities in vaccine accessibility that affected in particular equity-deserving individuals in Ontario at the time of the survey. However, because to the best of our knowledge there seems to be a very limited amount of literature on vaccine accesibility for marginalized groups in the context of the province, there is an ongoing need of future studies this topic in the different Health Regions of the province.

@@

However, the results also indicate that the place of habitation affected the odds of vaccination for certain equity-deserving groups (interaction term of Health Region and Race, @fig-models,B). Specifically, this held true in the case of individuals identifying as Indigenous or with mixed racial background in the West Health Region, Black individuals in the Central Health Region, and Arab/Middle Eastern individuals in the East Health Region [@fig-models]. For these individuals, vaccination odds were lower when compared to the Toronto Health Region (Supplementary Figure A-6). We indicate next some contributing factors that might help provide context to these results.

First, it is useful to dissagregate the data by LHINs, because most of the studies in the literature that have analyzed health in Ontario use the LHINs as the base of their analyses. Interestingly, for Indigenous and mixed individuals most of the observations in the survey for West Health Region came from the Hamilton Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs (**add table in Appendix**), whereas for Arab/Middle Eastern Individuals in the East Health Region the highest number of observations corresponded to the Champlain and Central East LHINs. Previous research has identified health disparities in these (mostly rural) regions, such as unequal distribution of primary care providers, increased mortality, and low pharmacist availability [@shah2019;@crighton2015;@timony2022].

### 2023-05-18 manuscript_v5/appendix.qmd
discussion, appendix updated

  clean_data %>%
  select(Race,Health_Region,first_dose) %>%
  tbl_strata(
    strata = Health_Region,
    .tbl_fun =
      ~ .x %>%
        tbl_summary(by = first_dose, missing = "no",percent="row") %>%
        add_n(),
    .header = "**{strata}**, N = {n}"
  )
```

@@

```{r,Central-HR-LHIN}
#| label: tbl-central-hr-lhin
#| tbl-cap: "Proportion of answers by LHIN (Central Health Region)"
#| message: false
#| echo: false

clean_data %>% subset (Health_Region=="Central")%>%
tbl_summary(include=c(LHIN),
            sort = all_categorical() ~ "frequency") %>% 
  bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")

@@

>>>>

  # clean_data %>%
  # select(Race,Health_Region,first_dose) %>%
  # tbl_strata(
  #   strata = Health_Region,
  #   .tbl_fun =
  #     ~ .x %>%
  #       tbl_summary(by = first_dose, missing = "no",percent="row") %>%
  #       add_n(),
  #   .header = "**{strata}**, N = {n}"
  # )%>%
  # bold_labels()%>%
  # as_kable_extra(booktabs = TRUE,linesep="")
```

@@

```{r,West-HR-Indigenous}
#| label: tbl-west-region-indigenous
#| tbl-cap: "Proportion of answers by LHIN in the West Health Region for Indigenous Individuals"
#| message: false
#| echo: false

clean_data %>%
  subset(Health_Region=="West" & Race=="Indigenous")%>%
  tbl_summary(by= first_dose,
              #missing="no",
              sort = list(everything() ~ "frequency"),
              include=c(LHIN))%>% 
  bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")

@@

```{r,West-HR-Mixed}
#| label: tbl-west-region-mixed
#| tbl-cap: "Proportion of answers by LHIN in the West Health Region for Mixed Individuals"
#| message: false
#| echo: false

clean_data %>%
  subset(Health_Region=="West" & Race=="Mixed")%>%
  tbl_summary(by= first_dose,
              #missing="no",
              sort = list(everything() ~ "frequency"),
              include=c(LHIN)) %>% 
  bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")

```{r,East-HR-Arab}
#| label: tbl-East-region-Arab
#| tbl-cap: "Proportion of answers by LHIN in the East Health Region for Arab/Middle Eastern Individuals"
#| message: false
#| echo: false

a1_rake %>%
  subset(Health_Region=="East" & Race=="Arab/Middle Eastern")%>%
  tbl_svysummary(by= first_dose,
              #missing="no",
              sort = list(everything() ~ "frequency"),
              include=c(LHIN))%>% 
  bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")

### 2023-05-18 manuscript_v5/appendix.tex
discussion, appendix updated

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata},
  pdfauthor={Ariel Mundo Ortiz1,2; Bouchra Nasri1,2,},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-central-hr-lhin}{}
\begin{table}
\caption{\label{tbl-central-hr-lhin}Proportion of answers by LHIN (Central Health Region) }\tabularnewline

\centering
\begin{tabular}{lc}
\toprule
\textbf{Characteristic} & \textbf{N = 1,455}\\
\midrule
\textbf{LHIN} & \\
\hspace{1em}Mississauga Halton & 495 (34\%)\\
\hspace{1em}Central & 428 (29\%)\\
\hspace{1em}Central West & 362 (25\%)\\
\hspace{1em}North Simcoe Muskoka & 170 (12\%)\\
\bottomrule
\multicolumn{2}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\end{tabular}
\end{table}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{placeins}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata},
  pdfauthor={Ariel Mundo Ortiz1,2; Bouchra Nasri1,2,},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-west-region-indigenous}{}
\begin{table}
\caption{\label{tbl-west-region-indigenous}Proportion of answers by LHIN in the West Health Region for Indigenous
Individuals }\tabularnewline

\centering
\begin{tabular}{lcc}
\toprule
\textbf{Characteristic} & \textbf{no}, N = 32 & \textbf{yes}, N = 42\\
\midrule
\textbf{LHIN} &  & \\
\hspace{1em}Hamilton Niagara Haldimand Brant (Hnhb) & 19 (59\%) & 9 (21\%)\\
\hspace{1em}South West & 4 (13\%) & 18 (43\%)\\
\hspace{1em}Waterloo Wellington & 7 (22\%) & 11 (26\%)\\
\hspace{1em}Erie St. Clair & 2 (6.3\%) & 4 (9.5\%)\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\end{tabular}
\end{table}

\hypertarget{tbl-west-region-mixed}{}
\begin{table}
\caption{\label{tbl-west-region-mixed}Proportion of answers by LHIN in the West Health Region for Mixed
Individuals }\tabularnewline

\centering
\begin{tabular}{lcc}
\toprule
\textbf{Characteristic} & \textbf{no}, N = 41 & \textbf{yes}, N = 76\\
\midrule
\textbf{LHIN} &  & \\
\hspace{1em}Waterloo Wellington & 14 (34\%) & 20 (26\%)\\
\hspace{1em}Hamilton Niagara Haldimand Brant (Hnhb) & 10 (24\%) & 22 (29\%)\\
\hspace{1em}South West & 11 (27\%) & 20 (26\%)\\
\hspace{1em}Erie St. Clair & 6 (15\%) & 14 (18\%)\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\end{tabular}
\end{table}

\hypertarget{tbl-East-region-Arab}{}
\begin{table}
\caption{\label{tbl-East-region-Arab}Proportion of answers by LHIN in the East Health Region for Arab/Middle
Eastern Individuals }\tabularnewline

\centering
\begin{tabular}{lcc}
\toprule
\textbf{Characteristic} & \textbf{no}, N = 19,716 & \textbf{yes}, N = 23,499\\
\midrule
\textbf{LHIN} &  & \\
\hspace{1em}Toronto \hspace{1em}Central & 0 (0\%) & 0 (0\%)\\
\hspace{1em}Champlain & 13,173 (67\%) & 19,825 (84\%)\\
\hspace{1em}Mississauga Halton & 0 (0\%) & 0 (0\%)\\
\hspace{1em}Hamilton Niagara Haldimand Brant (Hnhb) & 0 (0\%) & 0 (0\%)\\
Central & 0 (0\%) & 0 (0\%)\\
\hspace{1em}Waterloo Wellington & 0 (0\%) & 0 (0\%)\\
\hspace{1em}Central West & 0 (0\%) & 0 (0\%)\\
\hspace{1em}South West & 0 (0\%) & 0 (0\%)\\
\hspace{1em}Central East & 6,543 (33\%) & 3,674 (16\%)\\
\hspace{1em}Erie St. Clair & 0 (0\%) & 0 (0\%)\\
\hspace{1em}North Simcoe Muskoka & 0 (0\%) & 0 (0\%)\\
\hspace{1em}South East & 0 (0\%) & 0 (0\%)\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\end{tabular}
\end{table}

### 2023-05-18 manuscript_v5/main.tex
discussion, appendix updated

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto, and which ran between September 30th, 2021 and January
17th,2022. The survey collected socio-economic information from
participants (Table~\ref{tbl-descriptive-stats}), their location
(nearest municipality, as shown in Figure~\ref{fig-map}), the date of
access to the survey, and asked information on vaccination status by
using the question ``Have you received the first dose of the COVID
vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations (where each observation
corresponded to a unique respondent).

Preliminary analyses of the data included the removal of outliers
(\textbf{should we still do this? it's only 19 observations with income
\textgreater110k and household of 1, but we are not even using such
income bracket in the analysis because we re-grouped the data, and the
household size variable has 90\% missing rate}), of observations where
respondents did not provide answers in all the covariates of interest,
matching the city of each observations with its corresponding LHIN and
Health Region, and removing observations from areas with low
representation (107 observations corresponding to the North West and
North East Health Regions). After all the preliminary analyses indicated
above, the total number of observations used for analysis was 3,549
which included the East, Central, Toronto, and West Health Regions
covering the period between October 1st,2021 and December 12th, 2021.
The original dataset, clean dataset, and details on the data cleaning
process are described in detail in the
\href{https://github.com/aimundo/Fields_COVID-19/}{GitHub repository}
for this paper.

@@

Table~\ref{tbl-descriptive-stats} shows the characteristics of the data
from the Fields COVID-19 survey used for analysis. The sample contained
\textbf{6,236} observations, from which 24.8\% (1,547) corresponded to
individuals that reported not having received the first dose of the
vaccine. Vaccination rates ranged between 71-79\% across household
income brackets, age groups, Health Regions, and the months considered
in the survey. However, the highest vaccination rates in each category
were reported by individuals in the highest income bracket (79\%), those
between 16 and 34 years of age (77\%), individuals that lived in the
East Health Region (77\%), and during January of 2022 (78\%).
Differences were higher between racial/ethnic groups, where the higher
vaccination rate was reported by White/Caucasian individuals (84\%),
against vaccination rates between 63-66\% reported in the case of
Arab/Middle Eastern, Black, Indigenous, Latin American individuals, and
those that reported belonging to ``Other'' racial groups, which included
Southeast Asian, Filipino, West Asian, and minorities not identified
elsewhere.

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, boxrule=0pt, breakable, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto, and which ran between September 30th, 2021 and January
17th,2022. The survey collected socio-economic information from
participants (Table~\ref{tbl-descriptive-stats}), their location
(nearest municipality, as shown in Figure~\ref{fig-map}), the date of
access to the survey, and asked information on vaccination status by
using the question ``Have you received the first dose of the COVID
vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and
deployed the questions randomly, which resulted in \(\approx\) 84\% of
the observations having multiple missing answers or being incomplete.
Therefore, we selected 6,343 observations that were labeled as
``complete'' and that had complete answers for all covariates considered
in our analysis. Later, we matched the city of each observation with its
corresponding LHIN and Health Region, and removed observations from
areas with low representation (107 observations corresponding to the
North West and North East Health Regions). After all the preliminary
analyses indicated above, the total number of observations used for
analysis was 6,236 which included the East, Central, Toronto, and West
Health Regions covering the period between October 1st,2021 and January
17, 2022. The original dataset, clean dataset, and details on the data
cleaning process are described in detail in the
\href{https://github.com/aimundo/Fields_COVID-19/}{GitHub repository}
for this paper.

@@

Table~\ref{tbl-descriptive-stats} shows the characteristics of the data
from the Fields COVID-19 survey used for analysis. The sample contained
6,236 observations, from which 24.8\% (1,547) corresponded to
individuals that reported not having received the first dose of the
vaccine. Vaccination rates ranged between 71-79\% across household
income brackets, age groups, Health Regions, and the months considered
in the survey. However, the highest vaccination rates in each category
were reported by individuals in the highest income bracket (79\%), those
between 16 and 34 years of age (77\%), individuals that lived in the
East Health Region (77\%), and during January of 2022 (78\%).
Differences were higher between racial/ethnic groups, where the higher
vaccination rate was reported by White/Caucasian individuals (84\%),
against vaccination rates between 63-66\% reported in the case of
Arab/Middle Eastern, Black, Indigenous, Latin American individuals, and
those that reported belonging to ``Other'' racial groups, which included
Southeast Asian, Filipino, West Asian, and minorities not identified
elsewhere.
